---
title: "Semi-Supervised Text Classification of Thesis Metadata"
subtitle: "Classifying Ecology, Evolution, and Environment (EEE) Theses Using Tidymodels"
author: "Jason Pither, with help from Claude.ai (Sonnet 4.5)"
date: January 26, 2026
format:
  html:
    toc: true
    toc-location: left  
    toc-depth: 3
    code-fold: false
    code-tools: true
    theme: cosmo
    embed-resources: true
execute:
  cache: true
  warning: false
  message: false
---

# Introduction

## Overview

This document demonstrates a **semi-supervised machine learning approach** to classify thesis metadata into two categories:

1. **EEE (Ecology, Evolution, Environment)**: Theses related to ecological, evolutionary, or environmental sciences
2. **Other**: All other thesis topics

The classification is based solely on **text features** (thesis title and abstract), making the model generalizable to any thesis metadata regardless of available metadata fields like academic program.

We have metadata for 4,272 theses (published between 2022 and 2024 inclusive), obtained from the University of British Columbia's [public database](https://open.library.ubc.ca/cIRcle/collections/ubctheses).  

The metadata include a "Program" field that we can leverage for training purposes, and:

- Some theses are from Zoology or Botany programs → definitely EEE
- Most theses are from clearly non-EEE programs → definitely "other"
- Some theses are from ambiguous programs (Geography, Biology, etc.) where some may be EEE but we don't know which

The key challenge: **Some EEE theses are "hidden" in ambiguous programs**, and we don't know how many exist.

## Our Approach: Iterative Semi-Supervised Learning

We use an **active learning** strategy:

1. Train initial model on theses with reliable labels
2. Apply model to unlabeled theses
3. Manually review high-confidence predictions
4. Incorporate verified labels
5. Retrain model with expanded dataset
6. Repeat until satisfied

This allows us to progressively discover hidden EEE theses and improve model performance with each iteration.

# Setup

## Load Required Packages

```{r}
#| label: load-packages

# Core tidyverse for data manipulation
library(tidyverse)

# Tidymodels ecosystem for machine learning
library(tidymodels)

# Text preprocessing
library(textrecipes)
library(stopwords)

# File path management
library(here)

# Parallel processing
library(future)
library(future.apply)

# Model interpretation (for optional sections)
library(vip)
```

## Package Overview

- **tidyverse**: Data manipulation and visualization
- **tidymodels**: Unified framework for modeling (includes `parsnip`, `recipes`, `workflows`, `tune`, `yardstick`)
- **textrecipes**: Text-specific preprocessing steps for recipes
- **stopwords**: Common word removal
- **here**: Reproducible file paths
- **future**: Modern parallel processing
- **vip**: Variable importance plots

# Data Preparation

## Load Thesis Metadata

```{r}
#| label: load-data

thesis_data <- readr::read_csv(
  here::here("data", "processed_data", "comparator-theses", 
             "training-data", "ubc_thesis_data.csv")
)
```

Our dataset contains:

- **DOI**: Unique identifier for each thesis
- **Title**: Thesis title
- **Description**: Thesis abstract
- **Program (Theses)**: Academic program (used for labeling only, NOT as a model feature)

## Create Combined Text Feature

We combine title and abstract into a single text feature. This is what the model will learn from.

```{r}
#| label: combine-text

thesis_data <- thesis_data %>%
  dplyr::mutate(combined_text = paste(Title, Description, sep = " "))
```

**Why combine?** The model works with a single input, and both title and abstract contain informative terms. 

## Rename Program Field

Here we rename the Program field to be simpler.

```{r}
#| label: rename-program

thesis_data <- thesis_data %>%
  dplyr::rename(Program = "Program (Theses)")
```

# Creating Training Labels

## Understanding the Labeling Strategy

The academic program field is used **only for creating initial labels**, not as a model feature. This is crucial because the final model must work on thesis metadata that lack program information.

## Identify Program Categories

First, let's examine all unique programs to categorize them:

```{r}
#| label: explore-programs

# Get all unique programs, sorted alphabetically
unique_programs <- sort(unique(thesis_data$Program))

# If you'd like to see the full list of programs, then uncomment this:
# unique_programs
```

### Definite EEE Programs

These programs are, by definition, focused on ecology, evolution, or environment:

```{r}
#| label: positive-programs

positive_programs <- c("Zoology", "Botany")
```

### Definite Non-EEE Programs

These programs have no ecological/evolutionary/environmental content. We identify them systematically by keyword:

```{r}
#| label: negative-programs

# Extract programs by distinguishing keywords
business <- unique_programs[grep("Business", unique_programs)]
music <- unique_programs[grep("Music", unique_programs)]
engineering <- unique_programs[grep("Engineering", unique_programs)]
education <- unique_programs[grep("Education", unique_programs)]
arts <- unique_programs[grep("Art", unique_programs)]
human <- unique_programs[grep("Human", unique_programs)]
physics <- unique_programs[grep("Physics", unique_programs)]
pharma <- unique_programs[grep("Pharma", unique_programs)]
culture <- unique_programs[grep("Cultur", unique_programs)]
health <- unique_programs[grep("Health", unique_programs)]
medical <- unique_programs[grep("Medic", unique_programs)]
psych <- unique_programs[grep("Psychology", unique_programs)]

# Programs with "Studies" (except food-related)
studies <- unique_programs[grep("Studies", unique_programs)]
studies <- studies[-grep("Food", studies)]

# Programs with "Science" (except plant/earth/soil)
science <- unique_programs[grep("Science", unique_programs)]
science <- science[-sapply(c("Plant", "Earth", "Soil"), function(x) {grep(x, science)})]

# Other obvious non-EEE programs
others <- c(
  "Anthropology", "Architecture", "Astronomy", "Chemistry", "Classics",
  "Economics", "English", "French", "Geophysics", "History", "Journalism",
  "Kinesiology", "Mathematics", "Law", "Nursing", "Neuroscience", "Philosophy", 
  "Planning", "Sociology", "Statistics", "Surgery", "Theatre", 
  "Applied Animal Biology", "Social Work", "Children's Literature", 
  "Occupational and Environmental Hygiene", "Classical and Near Eastern Archaeology", 
  "Cell and Developmental Biology", "Teaching English as a Second Language",
  "Interdisciplinary Oncology", "Linguistics", 
  "Gender, Race, Sexuality and Social Justice", "Landscape Architecture", 
  "Measurement, Evaluation and Research Methodology", "Creative Writing", 
  "Biochemistry and Molecular Biology"
)

# Combine all negative programs
negative_programs <- c(
  business, music, engineering, education, arts, human, physics, pharma,
  culture, health, medical, psych, studies, science, others
)
```

### Ambiguous Programs

These programs might contain EEE theses but we can't be certain:

```{r}
#| label: check-ambiguous

# Programs that are neither definitively EEE nor definitively non-EEE
ambiguous_programs <- setdiff(unique_programs, c(negative_programs, positive_programs))
print(ambiguous_programs)
```

These ambiguous programs (Geography, Biology, Forestry, etc.) are where we expect to find hidden EEE theses through our iterative approach.

## Apply Labels

Now we label all theses based on their program:

```{r}
#| label: apply-labels

thesis_data <- thesis_data %>%
  dplyr::mutate(
    label_status = dplyr::case_when(
      Program %in% positive_programs ~ "positive",
      Program %in% negative_programs ~ "negative",
      TRUE ~ "unlabeled"
    ),
    category = dplyr::case_when(
      label_status == "positive" ~ "EEE",
      label_status == "negative" ~ "other",
      TRUE ~ NA_character_
    ),
    category = factor(category, levels = c("EEE", "other"))
  )
```

**What's happening here?**

- **label_status**: Tracks whether we're confident about the label
- **category**: The actual classification (EEE vs. other)
- **factor levels**: Explicitly set to ensure consistent ordering

## Review Initial Labeling

```{r}
#| label: review-labeling

# Summary by program and label status
labeling_summary <- thesis_data %>%
  dplyr::count(label_status, Program) %>%
  dplyr::arrange(label_status, desc(n))

print(labeling_summary)

# Overall label distribution
thesis_data %>%
  dplyr::count(label_status)
```

# Initial Model Training

## Extract Labeled Data

We start by training on **only the reliably labeled data**:

```{r}
#| label: extract-labeled

labeled_data <- thesis_data %>%
  dplyr::filter(!is.na(category)) %>%
  dplyr::select(DOI, combined_text, category, Title, Description)

# Check class balance
labeled_data %>% dplyr::count(category)
```

## Balance Classes

Machine learning models can be biased toward the majority class. We **downsample** to ensure equal representation:

```{r}
#| label: balance-classes

set.seed(123)  # For reproducibility

min_class_size <- min(table(labeled_data$category))

balanced_data <- labeled_data %>%
  dplyr::group_by(category) %>%
  dplyr::slice_sample(n = min_class_size) %>%
  dplyr::ungroup()
```

Training with `r nrow(balanced_data)` balanced examples: `r sum(balanced_data$category == "EEE")` EEE and `r sum(balanced_data$category == "other")` other.

**Why balance?** If we trained on imbalanced data (e.g., 90% "other"), the model might just predict "other" for everything and still achieve 90% accuracy.

## Create Train/Test Split

We hold out 20% of data to evaluate final model performance:

```{r}
#| label: train-test-split

data_split <- rsample::initial_split(balanced_data, prop = 0.80, strata = category)
train_data <- rsample::training(data_split)
test_data <- rsample::testing(data_split)
```

**Stratified split**: Maintains the same class proportions in both train and test sets.

## Create Cross-Validation Folds

For hyperparameter tuning, we'll use 5-fold cross-validation:

```{r}
#| label: create-cv-folds

cv_folds <- rsample::vfold_cv(train_data, v = 5, strata = category)
```

**What is cross-validation?** We split the training data into 5 parts. The model trains on 4 parts and tests on the 5th, rotating through all combinations. This gives us a robust estimate of model performance.

# Text Preprocessing Recipe

## Understanding the Recipe

A **recipe** defines all preprocessing steps applied to raw data before modeling. Think of it as a blueprint that will be applied consistently to training data, test data, and future predictions.

```{r}
#| label: create-recipe

text_recipe <- recipes::recipe(category ~ combined_text, data = train_data) %>%
  textrecipes::step_tokenize(combined_text) %>%
  textrecipes::step_stopwords(
    combined_text, 
    custom_stopword_source = stopwords::stopwords("en")
  ) %>%
  textrecipes::step_tokenfilter(
    combined_text, 
    min_times = 5, 
    max_tokens = 5000
  ) %>%
  textrecipes::step_tfidf(combined_text) %>%
  recipes::step_normalize(recipes::all_predictors())
```

## Recipe Steps Explained

### 1. Tokenization

**What it does:** Breaks text into individual words (tokens)

- Input: `"Species diversity varies across ecosystems"`
- Output: `["species", "diversity", "varies", "across", "ecosystems"]`

### 2. Stopword Removal

**What it does:** Removes common, uninformative words like "the", "and", "is"

**Why?** These words appear in all documents regardless of topic and don't help distinguish EEE from non-EEE theses.

### 3. Token Filtering

**What it does:** 

- `min_times = 5`: Removes tokens appearing in fewer than 5 documents (likely typos or rare jargon)
- `max_tokens = 5000`: Keeps only the 5000 most frequent remaining tokens

**Why?** Reduces noise and computational complexity while retaining informative terms.

### 4. TF-IDF Transformation

**What it does:** Converts tokens to numeric features using **Term Frequency-Inverse Document Frequency**

**TF-IDF explained:**

- **Term Frequency (TF)**: How often a word appears in a document
- **Inverse Document Frequency (IDF)**: How rare a word is across all documents
- **TF-IDF = TF × IDF**: Upweights words that are frequent in a document but rare overall

**Example:**

- Word "species" in an ecology thesis → high TF-IDF (frequent in this thesis, rare overall)
- Word "analysis" everywhere → low TF-IDF (common across all theses)

This gives each document a numeric vector where high values indicate distinctive, informative terms.

### 5. Normalization

**What it does:** Standardizes all features to mean=0, standard deviation=1

**Why?** Regularized models (which we'll use) are sensitive to feature scale. Normalization ensures all tokens contribute proportionally.

# Model Specification

## Define the Model

We use **elastic net regularized logistic regression**:

```{r}
#| label: model-specification

logistic_spec <- parsnip::logistic_reg(
  penalty = tune::tune(),
  mixture = tune::tune()
) %>%
  parsnip::set_engine("glmnet") %>%
  parsnip::set_mode("classification")
```

## Why This Model?

### Logistic Regression

Predicts the probability that a thesis is EEE:

$$P(EEE) = \frac{1}{1 + e^{-(\beta_0 + \beta_1 x_1 + ... + \beta_p x_p)}}$$

Where:

- $\beta_i$ are coefficients learned from data
- $x_i$ are TF-IDF features (one per token)

Each token gets a coefficient indicating its association with EEE (positive) or other (negative).

### Why Regularization?

With 5000 potential features and only a few hundred training examples, we have a **high-dimensional problem**. Without regularization, the model would:

- Overfit to training data
- Assign importance to noise
- Fail to generalize

**Elastic net regularization** adds a penalty that:

- Shrinks coefficients toward zero
- Performs automatic feature selection
- Improves generalization

### Hyperparameters to Tune

- **penalty (λ)**: Strength of regularization (higher = more shrinkage)
- **mixture (α)**: Balance between L1 (lasso) and L2 (ridge) penalties
  - α = 0: Ridge regression (shrinks all coefficients)
  - α = 1: Lasso regression (sets some coefficients to exactly 0)
  - 0 < α < 1: Elastic net (combination of both)

## Create Workflow

A **workflow** bundles the recipe and model together:

```{r}
#| label: create-workflow

thesis_workflow <- workflows::workflow() %>%
  workflows::add_recipe(text_recipe) %>%
  workflows::add_model(logistic_spec)
```

This ensures preprocessing and modeling steps are always applied together consistently.

# Hyperparameter Tuning

## Define Tuning Grid

We'll test 100 combinations of hyperparameters:

```{r}
#| label: tuning-grid

tune_grid <- dials::grid_regular(
  dials::penalty(range = c(-5, 0)),  # 10^-5 to 10^0
  dials::mixture(range = c(0, 1)),    # 0 (ridge) to 1 (lasso)
  levels = 10                          # 10 values each = 100 combinations
)
```

## Run Tuning with Cross-Validation

This tests all 100 combinations using 5-fold cross-validation:

```{r}
#| label: run-tuning

# Set up parallel processing
future::plan(multisession, workers = parallel::detectCores() - 1)

tune_results <- thesis_workflow %>%
  tune::tune_grid(
    resamples = cv_folds,
    grid = tune_grid,
    metrics = yardstick::metric_set(accuracy, roc_auc, sens, spec),
    control = tune::control_grid(save_pred = TRUE)
  )

# Return to sequential processing
future::plan(sequential)
```

**What's happening?**

For each of the 100 hyperparameter combinations:

1. Train on 4 folds
2. Test on the held-out fold
3. Repeat for all 5 folds
4. Average the performance metrics

This gives us 500 model fits total (100 combinations × 5 folds).

**Metrics tracked:**

- **Accuracy**: Overall correct classification rate
- **ROC AUC**: Quality of probability predictions (better for imbalanced data)
- **Sensitivity**: True positive rate (EEE recall)
- **Specificity**: True negative rate (other recall)

## Select Best Hyperparameters

```{r}
#| label: select-best

best_params <- tune_results %>%
  tune::select_best(metric = "roc_auc")
```

Best hyperparameters selected:

```{r}
#| label: show-best-params
#| echo: false

knitr::kable(best_params)
```

**Hyperparameter Tuning Results Explained**

This shows the optimal hyperparameters selected after testing 100 combinations (10 penalty values × 10 mixture values) through cross-validation.

The .config Column
`Preprocessor1_Model011` is just an internal identifier—it means this was the 11th hyperparameter combination tested out of 100.

The "penalty" value (`r best_params$penalty`) indicates the **regularization strength** (λ in the equations). 

- **Very low value** = minimal regularization
- This means the model is allowed to use most of the available features (tokens) without much shrinkage
- The model found that aggressive regularization wasn't needed—likely because of reasonably good signal-to-noise ratio in the text features

The "mixture" value (`r best_params$mixture`) indicates a balance between L1 (lasso) and L2 (ridge) penalties (α in the equations).

- Close to 0 = mostly ridge regression
- Ridge shrinks all coefficients proportionally but keeps all features
- Lasso (mixture = 1) would set some coefficients to exactly 0, performing feature selection
- At the observed value of (`r best_params$mixture`), the model uses **mostly ridge with a touch of lasso**—it shrinks coefficients but doesn't aggressively eliminate features

So, the current optimal model:
- Uses **light regularization** (small penalty)
- Keeps **most features** (low mixture = mostly ridge)
- Suggests the TF-IDF features are informative and not too noisy

With clear vocabulary differences between EEE and non-EEE theses:
- Many terms are genuinely predictive ("species", "ecological" vs "clinical", "algorithm")
- Heavy regularization would discard useful information
- You benefit from keeping many terms with small weights rather than selecting just a few

We use **ROC AUC** rather than accuracy because it evaluates performance across all possible classification thresholds, making it more robust.

## Finalize Workflow

```{r}
#| label: finalize-workflow

final_workflow <- thesis_workflow %>%
  tune::finalize_workflow(best_params)
```

## Train Final Model

Now we train using the best hyperparameters on all training data:

```{r}
#| label: train-final-model

final_fit <- final_workflow %>%
  parsnip::fit(data = balanced_data)
```

# Model Evaluation

## Predictions on Test Set

```{r}
#| label: test-predictions

test_predictions <- final_fit %>%
  predict(test_data) %>%
  dplyr::bind_cols(
    final_fit %>% predict(test_data, type = "prob")
  ) %>%
  dplyr::bind_cols(test_data)
```

**What we get:**

- `.pred_class`: Predicted category (EEE or other)
- `.pred_EEE`: Probability of being EEE
- `.pred_other`: Probability of being other

## Performance Metrics

```{r}
#| label: test-metrics

test_metrics <- test_predictions %>%
  yardstick::metrics(truth = category, estimate = .pred_class)
```

Test set performance:

```{r}
#| label: show-test-metrics
#| echo: false

knitr::kable(test_metrics, digits = 3)
```

## Confusion Matrix

```{r}
#| label: confusion-matrix

test_predictions %>%
  yardstick::conf_mat(truth = category, estimate = .pred_class)
```

**How to read:**

- **Rows**: True categories
- **Columns**: Predicted categories
- **Diagonal**: Correct predictions
- **Off-diagonal**: Errors

Clearly this model performed well, at least on the test set!

## Class-Specific Metrics

```{r}
#| label: class-metrics

sensitivity_val <- test_predictions %>% 
  yardstick::sens(truth = category, estimate = .pred_class) %>% 
  dplyr::pull(.estimate)

specificity_val <- test_predictions %>% 
  yardstick::spec(truth = category, estimate = .pred_class) %>% 
  dplyr::pull(.estimate)
```

- **Sensitivity (EEE recall)**: `r round(sensitivity_val, 3)` - Of all true EEE theses, what proportion did we correctly identify?
- **Specificity (Other recall)**: `r round(specificity_val, 3)` - Of all true "other" theses, what proportion did we correctly identify?

# Apply Model to Unlabeled Data

## Get Unlabeled Theses

```{r}
#| label: get-unlabeled

unlabeled_data <- thesis_data %>%
  dplyr::filter(label_status == "unlabeled") %>%
  dplyr::select(DOI, Program, Title, Description, combined_text)
```

Predicting on `r nrow(unlabeled_data)` unlabeled theses.

## Generate Predictions

```{r}
#| label: predict-unlabeled

unlabeled_predictions <- final_fit %>%
  predict(unlabeled_data, type = "prob") %>%
  dplyr::bind_cols(
    final_fit %>% predict(unlabeled_data, type = "class")
  ) %>%
  dplyr::bind_cols(unlabeled_data)
```

**Understanding the predictions:**

When we call `predict()` with `type = "class"`, the function assigns a category using a **0.5 probability threshold**:

- If P(EEE) > 0.5 → classified as "EEE"
- If P(EEE) ≤ 0.5 → classified as "other"

Using `type = "prob"` gives us the actual probabilities (`.pred_EEE` and `.pred_other`), which we can use to:

- Assess model confidence
- Apply custom thresholds (e.g., requiring P(EEE) > 0.8 for high-confidence classifications)
- Sort candidates by confidence for manual review

The 0.5 threshold assumes false positives and false negatives are equally costly. In our workflow, we review high-confidence predictions anyway, so this default works well. For production use on unlabeled data, you might adjust the threshold based on your tolerance for different types of errors.

## Identify High-Confidence EEE Candidates

```{r}
#| label: identify-candidates

eeb_candidates <- unlabeled_predictions %>%
  dplyr::filter(.pred_class == "EEE") %>%
  dplyr::arrange(desc(.pred_EEE)) %>%
  dplyr::select(DOI, Program, Title, Description, .pred_EEE, .pred_class) %>%
  dplyr::arrange(Program, Title)
```

Found `r nrow(eeb_candidates)` candidate EEE theses in unlabeled data.

## Export for Manual Review

```{r}
#| label: export-candidates
#| eval: false

readr::write_csv(
  eeb_candidates, 
  here::here("data", "processed_data", "comparator-theses", 
             "training-data", "ubc_eeb_candidate_theses_for_review_round1.csv")
)
```

**Next step:** Manually review this file and create a CSV with columns:

- `DOI`: Thesis identifier
- `verified_category`: Your manual classification ("EEE" or "other")

Save as `ubc_manually_reviewed_labels_round1.csv` for the next iteration.

# Iterative Refinement

## Round 1: Incorporate Manual Labels

After manually reviewing the first round of candidates, we incorporate those labels:

```{r}
#| label: round1-load-labels

manual_labels_round1 <- readr::read_csv(
  here::here("data", "processed_data", "comparator-theses", 
             "training-data", "ubc_manually_reviewed_labels_round1.csv")
) %>%
  dplyr::select(DOI, verified_category)

# Check review results
table(manual_labels_round1$verified_category)
```

So our manual assessment revealed `r table(manual_labels_round1$verified_category)[2]` theses that were categorized as "EEE", but determined to be "other".  

### Merge with Original Data

```{r}
#| label: round1-merge

thesis_data_v2 <- thesis_data %>%
  dplyr::left_join(manual_labels_round1, by = "DOI") %>%
  dplyr::mutate(
    # Update category: use verified label if available, otherwise keep original
    category = dplyr::coalesce(
      factor(verified_category, levels = c("EEE", "other")), 
      category
    ),
    # Update label_status: mark manually reviewed theses as labeled
    label_status = dplyr::if_else(
      !is.na(verified_category), 
      dplyr::if_else(verified_category == "EEE", "positive", "negative"),
      label_status
    )
  )
```

**What `coalesce()` does:** Uses the verified label if available, otherwise keeps the original label.

### Check Expansion

```{r}
#| label: round1-check-expansion

original_count <- sum(!is.na(thesis_data$category))
round1_count <- sum(!is.na(thesis_data_v2$category))
newly_added <- round1_count - original_count
```

**Label expansion summary:**

- Original labeled theses: `r original_count`
- After Round 1: `r round1_count`
- Newly added labels: `r newly_added`

### Save Updated Dataset

```{r}
#| label: round1-save
#| eval: false

readr::write_csv(
  thesis_data_v2, 
  here::here("data", "processed_data", "comparator-theses", 
             "training-data", "ubc_thesis_data_with_manual_labels_v1.csv")
)
```

## Round 1: Retrain Model

Now we retrain with the expanded labeled dataset:

```{r}
#| label: round1-retrain

# Extract labeled data
labeled_data_v2 <- thesis_data_v2 %>%
  dplyr::filter(!is.na(category)) %>%
  dplyr::select(DOI, combined_text, category, Title, Description)

# Balance classes
set.seed(1234)
min_class_size_v2 <- min(table(labeled_data_v2$category))

balanced_data_v2 <- labeled_data_v2 %>%
  dplyr::group_by(category) %>%
  dplyr::slice_sample(n = min_class_size_v2) %>%
  dplyr::ungroup()
```

Training with `r nrow(balanced_data_v2)` balanced examples: `r sum(balanced_data_v2$category == "EEE")` EEE and `r sum(balanced_data_v2$category == "other")` other.

```{r}
#| label: round1-split-folds

# Split and create folds
data_split_v2 <- rsample::initial_split(balanced_data_v2, prop = 0.80, strata = category)
train_data_v2 <- rsample::training(data_split_v2)
test_data_v2 <- rsample::testing(data_split_v2)
cv_folds_v2 <- rsample::vfold_cv(train_data_v2, v = 5, strata = category)

# Recipe
text_recipe_v2 <- recipes::recipe(category ~ combined_text, data = train_data_v2) %>%
  textrecipes::step_tokenize(combined_text) %>%
  textrecipes::step_stopwords(
    combined_text, 
    custom_stopword_source = stopwords::stopwords("en")
  ) %>%
  textrecipes::step_tokenfilter(combined_text, min_times = 5, max_tokens = 5000) %>%
  textrecipes::step_tfidf(combined_text) %>%
  recipes::step_normalize(recipes::all_predictors())

# Model specification
logistic_spec_v2 <- parsnip::logistic_reg(
  penalty = tune::tune(),
  mixture = tune::tune()
) %>%
  parsnip::set_engine("glmnet") %>%
  parsnip::set_mode("classification")

# Workflow
thesis_workflow_v2 <- workflows::workflow() %>%
  workflows::add_recipe(text_recipe_v2) %>%
  workflows::add_model(logistic_spec_v2)

# Tuning
tune_grid_v2 <- dials::grid_regular(
  dials::penalty(range = c(-5, 0)),
  dials::mixture(range = c(0, 1)),
  levels = 10
)

future::plan(multisession, workers = parallel::detectCores() - 1)

tune_results_v2 <- thesis_workflow_v2 %>%
  tune::tune_grid(
    resamples = cv_folds_v2,
    grid = tune_grid_v2,
    metrics = yardstick::metric_set(accuracy, roc_auc, sens, spec),
    control = tune::control_grid(save_pred = TRUE)
  )

future::plan(sequential)

# Select best and finalize
best_params_v2 <- tune_results_v2 %>%
  tune::select_best(metric = "roc_auc")

final_workflow_v2 <- thesis_workflow_v2 %>%
  tune::finalize_workflow(best_params_v2)

final_fit_v2 <- final_workflow_v2 %>%
  parsnip::fit(data = balanced_data_v2)
```

### Evaluate Round 1 Model

```{r}
#| label: round1-evaluate

test_predictions_v2 <- final_fit_v2 %>%
  predict(test_data_v2) %>%
  dplyr::bind_cols(
    final_fit_v2 %>% predict(test_data_v2, type = "prob")
  ) %>%
  dplyr::bind_cols(test_data_v2)

test_metrics_v2 <- test_predictions_v2 %>%
  yardstick::metrics(truth = category, estimate = .pred_class)
```

Test set performance (Round 1):

```{r}
#| label: round1-show-metrics
#| echo: false

knitr::kable(test_metrics_v2, digits = 3)
```

Confusion matrix (Round 1):

```{r}
#| label: round1-confusion
#| echo: false

test_predictions_v2 %>%
  yardstick::conf_mat(truth = category, estimate = .pred_class)
```
Again, perfect performance!

### Generate Round 2 Candidates

```{r}
#| label: round1-candidates

unlabeled_data_v2 <- thesis_data_v2 %>%
  dplyr::filter(label_status == "unlabeled") %>%
  dplyr::select(DOI, Program, Title, Description, combined_text)

unlabeled_predictions_v2 <- final_fit_v2 %>%
  predict(unlabeled_data_v2, type = "prob") %>%
  dplyr::bind_cols(
    final_fit_v2 %>% predict(unlabeled_data_v2, type = "class")
  ) %>%
  dplyr::bind_cols(unlabeled_data_v2)

eeb_candidates_v2 <- unlabeled_predictions_v2 %>%
  dplyr::filter(.pred_class == "EEE") %>%
  dplyr::arrange(desc(.pred_EEE)) %>%
  dplyr::select(DOI, Program, Title, Description, .pred_EEE, .pred_class) %>%
  dplyr::arrange(Program, Title)
```

Found `r nrow(eeb_candidates_v2)` candidate EEE theses for Round 2.

```{r}
#| label: round1-export
#| eval: false

readr::write_csv(
  eeb_candidates_v2, 
  here::here("data", "processed_data", "comparator-theses", 
             "training-data", "ubc_eeb_candidate_theses_for_review_round2.csv")
)
```

## Round 2: Manual Review and Retraining

```{r}
#| label: round2-process
set.seed(12345)

# Load Round 2 manual labels
manual_labels_round2 <- readr::read_csv(
  here::here("data", "processed_data", "comparator-theses", 
             "training-data", "ubc_manually_reviewed_labels_round2.csv")
) %>%
  dplyr::select(DOI, verified_category)

table(manual_labels_round2$verified_category)

# Merge into v3
thesis_data_v3 <- thesis_data_v2 %>%
  dplyr::select(-verified_category) %>%  # Remove old column to avoid conflict
  dplyr::left_join(manual_labels_round2, by = "DOI") %>%
  dplyr::mutate(
    category = dplyr::coalesce(
      factor(verified_category, levels = c("EEE", "other")), 
      category
    ),
    label_status = dplyr::if_else(
      !is.na(verified_category), 
      dplyr::if_else(verified_category == "EEE", "positive", "negative"),
      label_status
    )
  )

# Check progression
round2_count <- sum(!is.na(thesis_data_v3$category))
```

After Round 2: `r round2_count` labeled theses.

```{r}
#| label: round2-retrain-prep
#| echo: false

# Retrain
labeled_data_v3 <- thesis_data_v3 %>%
  dplyr::filter(!is.na(category)) %>%
  dplyr::select(DOI, combined_text, category, Title, Description)

min_class_size_v3 <- min(table(labeled_data_v3$category))

balanced_data_v3 <- labeled_data_v3 %>%
  dplyr::group_by(category) %>%
  dplyr::slice_sample(n = min_class_size_v3) %>%
  dplyr::ungroup()
```

Training with `r nrow(balanced_data_v3)` balanced examples.

```{r}
#| label: round2-split-continue
#| echo: false

data_split_v3 <- rsample::initial_split(balanced_data_v3, prop = 0.80, strata = category)
train_data_v3 <- rsample::training(data_split_v3)
test_data_v3 <- rsample::testing(data_split_v3)
cv_folds_v3 <- rsample::vfold_cv(train_data_v3, v = 5, strata = category)

text_recipe_v3 <- recipes::recipe(category ~ combined_text, data = train_data_v3) %>%
  textrecipes::step_tokenize(combined_text) %>%
  textrecipes::step_stopwords(
    combined_text, 
    custom_stopword_source = stopwords::stopwords("en")
  ) %>%
  textrecipes::step_tokenfilter(combined_text, min_times = 5, max_tokens = 5000) %>%
  textrecipes::step_tfidf(combined_text) %>%
  recipes::step_normalize(recipes::all_predictors())

logistic_spec_v3 <- parsnip::logistic_reg(
  penalty = tune::tune(),
  mixture = tune::tune()
) %>%
  parsnip::set_engine("glmnet") %>%
  parsnip::set_mode("classification")

thesis_workflow_v3 <- workflows::workflow() %>%
  workflows::add_recipe(text_recipe_v3) %>%
  workflows::add_model(logistic_spec_v3)

tune_grid_v3 <- dials::grid_regular(
  dials::penalty(range = c(-5, 0)),
  dials::mixture(range = c(0, 1)),
  levels = 10
)

future::plan(multisession, workers = parallel::detectCores() - 1)

tune_results_v3 <- thesis_workflow_v3 %>%
  tune::tune_grid(
    resamples = cv_folds_v3,
    grid = tune_grid_v3,
    metrics = yardstick::metric_set(accuracy, roc_auc, sens, spec),
    control = tune::control_grid(save_pred = TRUE)
  )

future::plan(sequential)

best_params_v3 <- tune_results_v3 %>%
  tune::select_best(metric = "roc_auc")

final_workflow_v3 <- thesis_workflow_v3 %>%
  tune::finalize_workflow(best_params_v3)

final_fit_v3 <- final_workflow_v3 %>%
  parsnip::fit(data = balanced_data_v3)

# Evaluate
test_predictions_v3 <- final_fit_v3 %>%
  predict(test_data_v3) %>%
  dplyr::bind_cols(
    final_fit_v3 %>% predict(test_data_v3, type = "prob")
  ) %>%
  dplyr::bind_cols(test_data_v3)

test_metrics_v3 <- test_predictions_v3 %>%
  yardstick::metrics(truth = category, estimate = .pred_class)
```

Test set performance (Round 2):

```{r}
#| label: round2-show-metrics
#| echo: false

knitr::kable(test_metrics_v3, digits = 3)
```

This time it's not perfect, but close!

```{r}
#| label: round2-candidates-gen
#| echo: false

# Generate Round 3 candidates
unlabeled_data_v3 <- thesis_data_v3 %>%
  dplyr::filter(label_status == "unlabeled") %>%
  dplyr::select(DOI, Program, Title, Description, combined_text)

unlabeled_predictions_v3 <- final_fit_v3 %>%
  predict(unlabeled_data_v3, type = "prob") %>%
  dplyr::bind_cols(
    final_fit_v3 %>% predict(unlabeled_data_v3, type = "class")
  ) %>%
  dplyr::bind_cols(unlabeled_data_v3)

eeb_candidates_v3 <- unlabeled_predictions_v3 %>%
  dplyr::filter(.pred_class == "EEE") %>%
  dplyr::arrange(desc(.pred_EEE)) %>%
  dplyr::select(DOI, Program, Title, Description, .pred_EEE, .pred_class) %>%
  dplyr::arrange(Program, Title)
```

Found `r nrow(eeb_candidates_v3)` candidate EEE theses for Round 3.

## Round 3: Manual Review and Retraining

```{r}
#| label: round3-process
set.seed(123456)
# Load Round 3 manual labels
manual_labels_round3 <- readr::read_csv(
  here::here("data", "processed_data", "comparator-theses", 
             "training-data", "ubc_manually_reviewed_labels_round3.csv")
) %>%
  dplyr::select(DOI, verified_category)

table(manual_labels_round3$verified_category)

# Merge into v4
thesis_data_v4 <- thesis_data_v3 %>%
  dplyr::select(-verified_category) %>%
  dplyr::left_join(manual_labels_round3, by = "DOI") %>%
  dplyr::mutate(
    category = dplyr::coalesce(
      factor(verified_category, levels = c("EEE", "other")), 
      category
    ),
    label_status = dplyr::if_else(
      !is.na(verified_category), 
      dplyr::if_else(verified_category == "EEE", "positive", "negative"),
      label_status
    )
  )

round3_count <- sum(!is.na(thesis_data_v4$category))
```

After Round 3: `r round3_count` labeled theses.

```{r}
#| label: round3-retrain-prep
#| echo: false

# Retrain
labeled_data_v4 <- thesis_data_v4 %>%
  dplyr::filter(!is.na(category)) %>%
  dplyr::select(DOI, combined_text, category, Title, Description)

min_class_size_v4 <- min(table(labeled_data_v4$category))

balanced_data_v4 <- labeled_data_v4 %>%
  dplyr::group_by(category) %>%
  dplyr::slice_sample(n = min_class_size_v4) %>%
  dplyr::ungroup()
```

Training with `r nrow(balanced_data_v4)` balanced examples.

```{r}
#| label: round3-split-continue
#| echo: false

data_split_v4 <- rsample::initial_split(balanced_data_v4, prop = 0.80, strata = category)
train_data_v4 <- rsample::training(data_split_v4)
test_data_v4 <- rsample::testing(data_split_v4)
cv_folds_v4 <- rsample::vfold_cv(train_data_v4, v = 5, strata = category)

text_recipe_v4 <- recipes::recipe(category ~ combined_text, data = train_data_v4) %>%
  textrecipes::step_tokenize(combined_text) %>%
  textrecipes::step_stopwords(
    combined_text, 
    custom_stopword_source = stopwords::stopwords("en")
  ) %>%
  textrecipes::step_tokenfilter(combined_text, min_times = 5, max_tokens = 5000) %>%
  textrecipes::step_tfidf(combined_text) %>%
  recipes::step_normalize(recipes::all_predictors())

logistic_spec_v4 <- parsnip::logistic_reg(
  penalty = tune::tune(),
  mixture = tune::tune()
) %>%
  parsnip::set_engine("glmnet") %>%
  parsnip::set_mode("classification")

thesis_workflow_v4 <- workflows::workflow() %>%
  workflows::add_recipe(text_recipe_v4) %>%
  workflows::add_model(logistic_spec_v4)

tune_grid_v4 <- dials::grid_regular(
  dials::penalty(range = c(-5, 0)),
  dials::mixture(range = c(0, 1)),
  levels = 10
)

future::plan(multisession, workers = parallel::detectCores() - 1)

tune_results_v4 <- thesis_workflow_v4 %>%
  tune::tune_grid(
    resamples = cv_folds_v4,
    grid = tune_grid_v4,
    metrics = yardstick::metric_set(accuracy, roc_auc, sens, spec),
    control = tune::control_grid(save_pred = TRUE)
  )

future::plan(sequential)

best_params_v4 <- tune_results_v4 %>%
  tune::select_best(metric = "roc_auc")

final_workflow_v4 <- thesis_workflow_v4 %>%
  tune::finalize_workflow(best_params_v4)

final_fit_v4 <- final_workflow_v4 %>%
  parsnip::fit(data = balanced_data_v4)

# Evaluate
test_predictions_v4 <- final_fit_v4 %>%
  predict(test_data_v4) %>%
  dplyr::bind_cols(
    final_fit_v4 %>% predict(test_data_v4, type = "prob")
  ) %>%
  dplyr::bind_cols(test_data_v4)

test_metrics_v4 <- test_predictions_v4 %>%
  yardstick::metrics(truth = category, estimate = .pred_class)
```

Test set performance (Round 3):

```{r}
#| label: round3-show-metrics
#| echo: false

knitr::kable(test_metrics_v4, digits = 3)
```

Again, not perfect, but close.


```{r}
#| label: round3-candidates-gen
#| echo: false

# Generate Round 4 candidates
unlabeled_data_v4 <- thesis_data_v4 %>%
  dplyr::filter(label_status == "unlabeled") %>%
  dplyr::select(DOI, Program, Title, Description, combined_text)

unlabeled_predictions_v4 <- final_fit_v4 %>%
  predict(unlabeled_data_v4, type = "prob") %>%
  dplyr::bind_cols(
    final_fit_v4 %>% predict(unlabeled_data_v4, type = "class")
  ) %>%
  dplyr::bind_cols(unlabeled_data_v4)

eeb_candidates_v4 <- unlabeled_predictions_v4 %>%
  dplyr::filter(.pred_class == "EEE") %>%
  dplyr::arrange(desc(.pred_EEE)) %>%
  dplyr::select(DOI, Program, Title, Description, .pred_EEE, .pred_class) %>%
  dplyr::arrange(Program, Title)
```

Found `r nrow(eeb_candidates_v4)` candidate EEE theses for Round 4.

## Round 4: Final Iteration

```{r}
#| label: round4-process
set.seed(1234)
# Load Round 4 manual labels
manual_labels_round4 <- readr::read_csv(
  here::here("data", "processed_data", "comparator-theses", 
             "training-data", "ubc_manually_reviewed_labels_round4.csv")
) %>%
  dplyr::select(DOI, verified_category)

table(manual_labels_round4$verified_category)

# Merge into v5 (final)
thesis_data_v5 <- thesis_data_v4 %>%
  dplyr::select(-verified_category) %>%
  dplyr::left_join(manual_labels_round4, by = "DOI") %>%
  dplyr::mutate(
    category = dplyr::coalesce(
      factor(verified_category, levels = c("EEE", "other")), 
      category
    ),
    label_status = dplyr::if_else(
      !is.na(verified_category), 
      dplyr::if_else(verified_category == "EEE", "positive", "negative"),
      label_status
    )
  )
```

**Label progression summary:**

```{r}
#| label: progression-summary
#| echo: false

progression <- tibble::tibble(
  Iteration = c("Original (v1)", "Round 1 (v2)", "Round 2 (v3)", "Round 3 (v4)", "Round 4 (v5)"),
  `Labeled Theses` = c(
    sum(!is.na(thesis_data$category)),
    sum(!is.na(thesis_data_v2$category)),
    sum(!is.na(thesis_data_v3$category)),
    sum(!is.na(thesis_data_v4$category)),
    sum(!is.na(thesis_data_v5$category))
  )
)

knitr::kable(progression)
```

```{r}
#| label: final-training-prep
#| echo: false

# Final training
labeled_data_v5 <- thesis_data_v5 %>%
  dplyr::filter(!is.na(category)) %>%
  dplyr::select(DOI, combined_text, category, Title, Description)

min_class_size_v5 <- min(table(labeled_data_v5$category))

balanced_data_v5 <- labeled_data_v5 %>%
  dplyr::group_by(category) %>%
  dplyr::slice_sample(n = min_class_size_v5) %>%
  dplyr::ungroup()
```

Training with `r nrow(balanced_data_v5)` balanced examples: `r sum(balanced_data_v5$category == "EEE")` EEE and `r sum(balanced_data_v5$category == "other")` other.

```{r}
#| label: final-split
#| echo: false

data_split_v5 <- rsample::initial_split(balanced_data_v5, prop = 0.80, strata = category)
train_data_v5 <- rsample::training(data_split_v5)
test_data_v5 <- rsample::testing(data_split_v5)
cv_folds_v5 <- rsample::vfold_cv(train_data_v5, v = 5, strata = category)

text_recipe_v5 <- recipes::recipe(category ~ combined_text, data = train_data_v5) %>%
  textrecipes::step_tokenize(combined_text) %>%
  textrecipes::step_stopwords(
    combined_text, 
    custom_stopword_source = stopwords::stopwords("en")
  ) %>%
  textrecipes::step_tokenfilter(combined_text, min_times = 5, max_tokens = 5000) %>%
  textrecipes::step_tfidf(combined_text) %>%
  recipes::step_normalize(recipes::all_predictors())

logistic_spec_v5 <- parsnip::logistic_reg(
  penalty = tune::tune(),
  mixture = tune::tune()
) %>%
  parsnip::set_engine("glmnet") %>%
  parsnip::set_mode("classification")

thesis_workflow_v5 <- workflows::workflow() %>%
  workflows::add_recipe(text_recipe_v5) %>%
  workflows::add_model(logistic_spec_v5)

tune_grid_v5 <- dials::grid_regular(
  dials::penalty(range = c(-5, 0)),
  dials::mixture(range = c(0, 1)),
  levels = 10
)

future::plan(multisession, workers = parallel::detectCores() - 1)

tune_results_v5 <- thesis_workflow_v5 %>%
  tune::tune_grid(
    resamples = cv_folds_v5,
    grid = tune_grid_v5,
    metrics = yardstick::metric_set(accuracy, roc_auc, sens, spec),
    control = tune::control_grid(save_pred = TRUE)
  )

future::plan(sequential)

best_params_v5 <- tune_results_v5 %>%
  tune::select_best(metric = "roc_auc")
```

Best hyperparameters (Round 4):

```{r}
#| label: show-best-params-v5
#| echo: false

knitr::kable(best_params_v5)
```

```{r}
#| label: finalize-fit-v5
#| echo: false

final_workflow_v5 <- thesis_workflow_v5 %>%
  tune::finalize_workflow(best_params_v5)

final_fit_v5 <- final_workflow_v5 %>%
  parsnip::fit(data = balanced_data_v5)
```

## Final Model Performance

```{r}
#| label: final-performance

test_predictions_v5 <- final_fit_v5 %>%
  predict(test_data_v5) %>%
  dplyr::bind_cols(
    final_fit_v5 %>% predict(test_data_v5, type = "prob")
  ) %>%
  dplyr::bind_cols(test_data_v5)

test_metrics_v5 <- test_predictions_v5 %>%
  yardstick::metrics(truth = category, estimate = .pred_class)

sensitivity_v5 <- test_predictions_v5 %>% 
  yardstick::sens(truth = category, estimate = .pred_class) %>% 
  dplyr::pull(.estimate)

specificity_v5 <- test_predictions_v5 %>% 
  yardstick::spec(truth = category, estimate = .pred_class) %>% 
  dplyr::pull(.estimate)
```

Final test set performance:

```{r}
#| label: show-final-metrics
#| echo: false

knitr::kable(test_metrics_v5, digits = 3)
```

Final confusion matrix:

```{r}
#| label: final-confusion
#| echo: false

test_predictions_v5 %>%
  yardstick::conf_mat(truth = category, estimate = .pred_class)
```

- **Sensitivity (EEE recall)**: `r round(sensitivity_v5, 3)`
- **Specificity (Other recall)**: `r round(specificity_v5, 3)`

## Remaining Unlabeled Theses

```{r}
#| label: final-unlabeled

unlabeled_data_v5 <- thesis_data_v5 %>%
  dplyr::filter(label_status == "unlabeled") %>%
  dplyr::select(DOI, Program, Title, Description, combined_text)

unlabeled_predictions_v5 <- final_fit_v5 %>%
  predict(unlabeled_data_v5, type = "prob") %>%
  dplyr::bind_cols(
    final_fit_v5 %>% predict(unlabeled_data_v5, type = "class")
  ) %>%
  dplyr::bind_cols(unlabeled_data_v5)

eeb_candidates_v5 <- unlabeled_predictions_v5 %>%
  dplyr::filter(.pred_class == "EEE") %>%
  dplyr::arrange(desc(.pred_EEE)) %>%
  dplyr::select(DOI, Program, Title, Description, .pred_EEE, .pred_class) %>%
  dplyr::arrange(Program, Title)
```

Remaining unlabeled theses: `r nrow(unlabeled_data_v5)`. Found `r nrow(eeb_candidates_v5)` candidate EEE theses in remaining data.

With only `r nrow(unlabeled_data_v5)` unlabeled theses and `r nrow(eeb_candidates_v5)` candidates remaining, we've reached a natural stopping point. The iterative process has successfully labeled `r round((1 - nrow(unlabeled_data_v5)/nrow(thesis_data)) * 100, 1)`% of the dataset.

# Save Final Model

## Export Production Model

```{r}
#| label: save-model
#| eval: false

saveRDS(
  final_fit_v5, 
  here::here("data", "processed_data", "comparator-theses", 
             "training-data", "eee_text_classifier.rds")
)
```

## Save Model Metadata

```{r}
#| label: save-metadata
#| eval: false

model_info <- list(
  model_file = "eee_text_classifier.rds",
  training_date = Sys.Date(),
  training_size = nrow(balanced_data_v5),
  n_eeb = sum(balanced_data_v5$category == "EEE"),
  n_other = sum(balanced_data_v5$category == "other"),
  test_accuracy = test_metrics_v5 %>%
    dplyr::filter(.metric == "accuracy") %>%
    dplyr::pull(.estimate),
  features = "Title + Description",
  note = "Model works on any thesis with Title & Description"
)

saveRDS(
  model_info, 
  here::here("data", "processed_data", "comparator-theses", 
             "training-data", "eee_text_classifier_model_info.rds")
)
```

# Using the Model on New Data

## Loading and Applying the Model

```{r}
#| label: example-usage
#| eval: false

# Load the saved model
model <- readRDS(
  here::here("data", "processed_data", "comparator-theses", 
             "training-data", "eee_text_classifier.rds")
)

# Prepare new data (must have 'Title' and 'Description' columns)

# **Note**
# UBC's use of the "Description" field name is unusual; most others
# will likely use "Abstract", and all thesis metadata downloaded from
# the Library Archives of Canada use "Abstract", so:

new_theses <- readr::read_csv("path/to/new_theses.csv") %>%
  dplyr::mutate(combined_text = paste(Title, Abstract, sep = " "))

# Generate predictions
predictions <- model %>%
  predict(new_theses, type = "prob") %>%
  dplyr::bind_cols(model %>% predict(new_theses, type = "class")) %>%
  dplyr::bind_cols(new_theses)

# Filter high-confidence EEE predictions
eee_predictions <- predictions %>%
  dplyr::filter(.pred_EEE > 0.80) %>%
  dplyr::arrange(desc(.pred_EEE))
```

# Find publications from thesis authors

In this step, we first identify the author ("Creator") of the thesis.  Later, a different script can use that author name to find publications that could be associated with the thesis research.

```{r}
#| label: find-author
#| eval: false

# TO BE COMPLETED


```


## Understanding Classification Thresholds

The default classification threshold is **0.5**: 

- P(EEE) > 0.5 → classify as "EEE"
- P(EEE) ≤ 0.5 → classify as "other"

You can adjust this threshold based on your use case:

```{r}
#| label: custom-threshold
#| eval: false

# For high precision (fewer false positives)
high_confidence_eee <- predictions %>%
  dplyr::mutate(
    custom_class = dplyr::if_else(.pred_EEE >= 0.8, "EEE", "other")
  )

# For high recall (catch more true EEE, accept false positives)
liberal_eee <- predictions %>%
  dplyr::mutate(
    custom_class = dplyr::if_else(.pred_EEE >= 0.4, "EEE", "other")
  )
```

# Summary and Key Takeaways

Starting with **560 unlabeled theses** from ambiguous programs, through 4 rounds of iterative manual review, we:

1. Labeled **395 additional theses** through active learning
2. Achieved **100% test accuracy** on the final model
3. Reduced unlabeled theses to **165** (only 3.9% of dataset)
4. Created a production-ready classifier requiring only title and abstract

## Lessons

### 1. Semi-Supervised Learning Works

When you have:

- Reliable positive examples (Zoology/Botany)
- Reliable negative examples (clearly non-EEE programs)
- Unlabeled data containing hidden positives

**Active learning** efficiently discovers hidden examples through iterative human-in-the-loop refinement.

### 2. Text Features Are Powerful

Using only **title and abstract**, the model learned to distinguish:

- **EEE terms**: "species", "ecological", "biodiversity", "conservation"
- **Non-EEE terms**: "cancer", "clinical", "treatment", "algorithm"

This vocabulary-based approach generalizes well beyond program names.

### 3. Minimal Regularization Was Sufficient

The optimal hyperparameters (penalty = 0.00001, mixture = 0) reveal that:

- **Very light regularization** was best - almost unpenalized
- **Pure ridge regression** (no feature selection) outperformed elastic net
- This suggests clean, informative text features with good signal-to-noise

**Why this worked:** EEE and non-EEE theses have distinctive vocabularies, and the TF-IDF preprocessing effectively captured these differences. The token filtering step (removing terms appearing in <5 documents) eliminated enough noise that aggressive regularization wasn't necessary.

This meant that the classification task had clear boundaries and the model could confidently use information from many terms rather than having to be heavily constrained.

### 4. Iteration Yields Diminishing Returns

- **Round 1**: 264 candidates → 208 EEE (79%)
- **Round 2**: 103 candidates → 66 EEE (64%)
- **Round 3**: 19 candidates → 10 EEE (53%)
- **Round 4**: 9 candidates → 6 EEE (67%)

As unlabeled data shrinks, the model becomes better calibrated but finds fewer new examples.

## Model Limitations

1. **Requires quality text**: Thin or poorly-written abstracts reduce performance
2. **Domain-specific**: Trained on academic theses; may not generalize to other document types
3. **Language-dependent**: English stopwords and tokenization; won't work on other languages

## Future Improvements

### For Production Use

1. **Ensemble models**: Combine multiple algorithms for robustness
2. **Bigrams/trigrams**: Capture phrases like "natural selection"
3. **Domain dictionaries**: Custom lists of EEE vs. non-EEE terms
4. **Confidence calibration**: Ensure probabilities match empirical frequencies

### For Advanced Users

The repetitive training code could be refactored into functions:

```{r}
#| label: refactoring-example
#| eval: false

# Example function structure
train_iteration <- function(data, iteration_name, seed = 1234) {
  # Balance classes
  # Split data
  # Create recipe
  # Tune hyperparameters
  # Train final model
  # Return list(model, metrics, predictions)
}

# Usage
results <- list()
results$round1 <- train_iteration(thesis_data_v2, "round1")
results$round2 <- train_iteration(thesis_data_v3, "round2")
```

This would reduce code duplication while maintaining clarity for beginners.

# Appendix: Optional Analyses

## Model Interpretation: Top Predictive Terms

To understand which words drive predictions:

```{r}
#| label: top-terms
#| eval: false

top_terms <- final_fit_v5 %>%
  workflows::extract_fit_parsnip() %>%
  vip::vi(lambda = best_params_v5$penalty) %>%
  dplyr::mutate(
    Direction = dplyr::if_else(Sign == "POS", "Other", "EEE"),
    abs_importance = abs(Importance)
  ) %>%
  dplyr::group_by(Direction) %>%
  dplyr::slice_max(abs_importance, n = 20) %>%
  dplyr::ungroup()

# View top EEE predictors
top_terms %>%
  dplyr::filter(Direction == "EEE") %>%
  dplyr::select(Variable, Importance) %>%
  print(n = 20)

# View top "other" predictors
top_terms %>%
  dplyr::filter(Direction == "Other") %>%
  dplyr::select(Variable, Importance) %>%
  print(n = 20)
```

### Visualize Top Terms

```{r}
#| label: visualize-terms
#| eval: false

top_terms %>%
  ggplot2::ggplot(
    ggplot2::aes(
      x = Importance, 
      y = reorder(Variable, Importance), 
      fill = Direction
    )
  ) +
  ggplot2::geom_col() +
  ggplot2::facet_wrap(~Direction, scales = "free_y") +
  ggplot2::labs(
    title = "Top 20 Predictive Terms by Category",
    y = NULL,
    x = "Coefficient"
  ) +
  ggplot2::theme_minimal()

ggplot2::ggsave("top_predictive_terms.png", width = 12, height = 8)
```

This helps validate that the model learned sensible patterns and can guide manual review by highlighting unexpected associations.

---

## Session Info

```{r}
#| label: session-info

sessionInfo()
```
