---
title: "Semi-Supervised Text Classification of Thesis Metadata (v2)"
subtitle: "Keyword-Seeded EEE Classification Using Multi-Institutional Training Data"
author: "Jason Pither, with help from Claude (Sonnet 4.6)"
date: February 2026
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: false
    code-tools: true
    theme: cosmo
    embed-resources: true
    df-print: paged
execute:
  cache: true
  warning: false
  message: false
---

# Introduction

## Motivation for v2

The original classifier (v1, `02_thesis_classification_model_training.qmd`) was trained
exclusively on UBC thesis data, using Zoology and Botany program membership as the source
of positive (EEE) labels. This created two problems:

1. **Narrow positive labels**: The model learned "Zoology/Botany vocabulary" rather
   than generalizable EEE signal, since only those two programs were used as anchors.

2. **Dissimilar negatives**: The negative training examples (Business, Medicine, Physics,
   etc.) were lexically very distant from EEE, so the model never learned to distinguish
   EEE from *near-neighbour* fields — geology, physical geography, and environmental
   social science — which produced the false positives observed in practice.

## Approach

This script replaces program-based labeling with **keyword-based labeling** applied
across all seven comparator institutions. The key changes:

1. **Multi-institutional training data**: All clean CSV files are pooled, giving broader
   vocabulary coverage than UBC alone.

2. **Keyword-seeded positives**: Theses matching 2 or more EEE-specific keyword patterns
   are assigned a confident EEE label. The 2-keyword threshold substantially reduces
   false positives relative to single-keyword matching.

3. **Explicit near-neighbour negatives**: Geology and environmental-social-science theses
   are deliberately included in the negative training set, targeting the specific false
   positive categories observed in v1.

4. **Same semi-supervised workflow**: The iterative active learning approach from v1 is
   retained. The keyword labels are the starting point; manual review of model predictions
   on ambiguous theses refines the model across rounds.

**NOTE**: For our project goals, we are more concerned about false positives than false negatives; we want the theses that we retain for downstream purposes to be confidently EEE-related theses, and it's not a big issue if some EEE theses are missed (assuming we don't end up being limited in number of candidate theses). 

## What remains the same as v1

- Elastic net logistic regression on TF-IDF features (tidymodels + textrecipes)
- Class balancing via downsampling
- Stratified train/test split and 5-fold cross-validation
- Iterative semi-supervised refinement with manual review

## Data

The 7 clean CSV files in `data/processed_data/comparator-theses/clean/` are used.
Each contains: `institution`, `title`, `abstract`, `author`, `year`, `program`.

```{r}
#| label: load-packages

library(tidyverse)
library(tidymodels)
library(textrecipes)
library(stopwords)
library(here)
library(future)
library(future.apply)
library(vip)
```

---

# Data Preparation

## Load and Pool All Clean CSVs

```{r}
#| label: load-data

# Load all clean CSVs and pool into one data frame
thesis_data <- list.files(
  here::here("data", "processed_data", "comparator-theses", "clean"),
  pattern = "\\.csv$",
  full.names = TRUE
) |>
  purrr::map(readr::read_csv, show_col_types = FALSE) |>
  dplyr::bind_rows()

cat("Total records:", nrow(thesis_data), "\n")
cat("Institutions:", paste(sort(unique(thesis_data$institution)), collapse = ", "), "\n")
```

## Create Combined Text Feature

```{r}
#| label: combine-text

thesis_data <- thesis_data |>
  dplyr::mutate(
    abstract = dplyr::if_else(is.na(abstract), "", abstract),
    combined_text = paste(title, abstract, sep = " ")
  )
```

---

# Keyword Labeling Strategy

## Why Keywords Instead of Program Names?

Using program names to generate training labels (as in v1) ties the model to the
vocabulary of specific departments at one institution. Keywords are more portable:
a thesis about "population dynamics of benthic macroinvertebrates" is EEE regardless
of whether it was filed under Zoology, Biology, Natural Resources, or an interdisciplinary
program.

**Important caveat**: keyword-labeled positives are not perfect ground truth. The
2-keyword threshold and compound patterns (e.g., `\bclimate change\b.{0,80}\bspecies\b`)
substantially improve precision, but some false positives remain. The manual review
steps later in this script address this.

## EEE Keyword Patterns

Patterns are grouped by E-E-E domain. Compound patterns (e.g., `climate change` +
`species`) are used where single terms are too ambiguous. Note `\bevolutionary\b` alone
is excluded because it fires on evolutionary algorithms and stellar evolution codes;
instead we require a biological qualifier after `evolutionary`.

```{r}
#| label: eee-keywords

eee_patterns <- list(
  # --- Ecology core ---
  ecology         = r"((?<!social )(?<!socio.)\becolog(?!ical (tea|museum|system)))",
  ecosystem       = r"(\becosystem\b)",
  habitat         = r"(\bhabitat\b)",
  biodiversity    = r"(\bbiodiversity\b)",
  species_richness = r"(\bspecies richness\b)",
  species_diversity = r"(\bspecies diversity\b)",
  food_web        = r"(\bfood web\b)",
  trophic         = r"(\btrophic\b)",
  riparian        = r"(\briparian\b)",
  benthic         = r"(\bbenthic\b)",
  macroinvert     = r"(\bmacroinvertebrate)",
  zooplankton     = r"(\bzooplankton\b)",
  phytoplankton   = r"(\bphytoplankton\b)",
  nutrient_cycling = r"(\bnutrient cycling\b)",
  biome           = r"(\bbiome\b)",
  pop_dynamics    = r"(\bpopulation dynamics\b)",

  # --- Evolution core (biological context required) ---
  evol_biology    = r"(\bevolutionary (biol|ecol|genet|physiol|histor|rate|dynamic|adapt|change))",
  phylogen        = r"(\bphylogen)",
  biogeograph     = r"(\bbiogeograph)",
  speciation      = r"(\bspeciation\b)",
  nat_selection   = r"(\bnatural selection\b)",
  sex_selection   = r"(\bsexual selection\b)",
  genetic_drift   = r"(\bgenetic drift\b)",
  pop_genetics    = r"(\bpopulation genetic)",
  mol_evol        = r"(\bmolecular evol)",
  coevol          = r"(\bcoevolut)",
  adapt_radiation = r"(\badaptive radiation\b)",

  # --- Environment (EEE-framed; compound patterns reduce ambiguity) ---
  climate_species = r"(\bclimate change\b.{0,80}\bspecies\b)",
  climate_ecol    = r"(\bclimate change\b.{0,80}\becolog)",
  landscape_ecol  = r"(\blandscape ecol)",
  hab_frag        = r"(\bhabitat fragmentation\b)",
  invasive_sp     = r"(\binvasive species\b)",
  conserv_biol    = r"(\bconservation biol)",
  restor_ecol     = r"(\brestoration ecol)",
  bioremediat     = r"(\bbioremediat)",
  wildlife_hab    = r"(\bwildlife\b.{0,40}\bhabitat\b)",
  wildlife_pop    = r"(\bwildlife\b.{0,40}\bpopulation\b)",
  wetland_ecol    = r"(\bwetland\b.{0,60}\becolog)",

  # --- Taxonomic / organismal in EEE context ---
  entomol         = r"(\bentomolog)",
  ornithol        = r"(\bornitholog)",
  ichthyol        = r"(\bichthyolog)",
  herpetol        = r"(\bherpetolog)",
  mycol           = r"(\bmycolog)",
  amphibian_pop   = r"(\bamphibian\b.{0,50}\bpopulation\b)",
  bird_species    = r"(\bbird\b.{0,50}\bspecies\b)",
  fish_pop        = r"(\bfish\b.{0,50}\bpopulation\b)",
  plant_communit  = r"(\bplant communit)",
  veg_communit    = r"(\bvegetation\b.{0,50}\bcommunit)",
  grassland_ecol  = r"(\bgrassland\b.{0,60}\bspecies\b)",
  forest_biodiv   = r"(\bforest\b.{0,60}\bbiodiversity\b)",
  marine_ecol     = r"(\bmarine\b.{0,50}\becolog)",
  freshwater_ecol = r"(\bfreshwater\b.{0,50}\becolog)"
)
```

## Near-Neighbour Negative Keyword Patterns

These patterns target the specific false-positive categories observed in v1: geology
and environmental social science. Theses matching these patterns (and no EEE patterns)
are explicitly included in the negative training class, forcing the model to learn the
distinction.

```{r}
#| label: neighbour-neg-keywords

# Geology: physical/structural geology terms unlikely in biological EEE
geology_patterns <- c(
  r"(\btectonic)", r"(\bseismic\b)", r"(\bstratigraph)",
  r"(\blitholog)", r"(\bgeomorph)", r"(\bsedimentary\b)",
  r"(\bmetamorphic\b)", r"(\bvolcan)", r"(\bmagma\b)",
  r"(\bigneous\b)", r"(\bgeochemistry\b)", r"(\bgeochronolog)",
  r"(\bhydrogeolog)", r"(\bgeophysics\b)", r"(\bpaleontolog)",
  r"(\bore deposit)",
  r"(\brare earth\b)",  # rare earth element mining/geochemistry
  r"(\blithium\b)",     # lithium mineral resources
  r"(\bcoal\b)"         # coal geology / combustion byproducts
)

# Environmental social science: policy/governance framing without biological content
env_soc_patterns <- c(
  r"(\benvironmental policy\b)", r"(\benvironmental governance\b)",
  r"(\benvironmental justice\b)", r"(\benvironmental management\b)",
  r"(\benvironmental law\b)", r"(\bclimate policy\b)",
  r"(\bclimate governance\b)", r"(\benergy policy\b)",
  r"(\bgreen economy\b)", r"(\bcarbon tax\b)"
)
```

## General Non-EEE Keyword Patterns

These terms reliably indicate human medicine, social science, education, or technology
research that is not EEE. They are kept as a separate group (rather than folded into
`env_soc_patterns`) so their contribution can be tracked independently. Given the
preference for low false positives over low false negatives, theses flagged by any of
these terms (with no EEE keyword matches) are assigned to the negative training pool.

Note: bare `policy` is deliberately excluded — "conservation policy", "wildlife policy",
and "fisheries management policy" appear in genuine EEE abstracts. The specific policy
phrases that are unambiguously non-EEE (climate policy, energy policy, etc.) are already
captured in `env_soc_patterns` above.

```{r}
#| label: general-nonEEE-keywords

general_nonEEE_patterns <- c(
  r"(\bvitamin\b)",        # human/clinical nutrition; rarely relevant to EEE
  r"(\bsoftware\b)",       # computer science / engineering
  r"(\bbusiness\b)",       # business / management
  r"(\bsocial justice\b)", # social sciences
  r"(\bnarrative)",        # narrative/narratives: qualitative social science
  r"(\beducat)",           # education / educational (env. education is soc sci not EEE)
  r"(\blanguage\b)",       # linguistics / qualitative research
  r"(\literacy\b)",        # qualitative research
  r"(\bmotiv\b)",          # for the word "motivation" or "motivated" which appears in qualitative research
  r"(\bfairness\b)",       # for the word "fairness" which is not typical of EEE research
  r"(\bpolitical\b)"       # political science (incl. political ecology, which is soc sci)
)
```

## Apply Keyword Labels

```{r}
#| label: apply-keywords

# Helper: count how many EEE patterns match in a text string
count_eee_matches <- function(text) {
  text_lower <- stringr::str_to_lower(text)
  sum(purrr::map_lgl(eee_patterns, ~ stringr::str_detect(text_lower, .x)))
}

# Helper: does text match any pattern in a list?
match_any <- function(text, patterns) {
  text_lower <- stringr::str_to_lower(text)
  any(purrr::map_lgl(patterns, ~ stringr::str_detect(text_lower, .x)))
}

# Apply to all theses (may take a minute)
thesis_data <- thesis_data |>
  dplyr::mutate(
    n_eee_kw  = purrr::map_int(combined_text, count_eee_matches),
    is_geo    = purrr::map_lgl(combined_text, ~ match_any(.x, geology_patterns)),
    is_envsoc = purrr::map_lgl(combined_text, ~ match_any(.x, env_soc_patterns)),
    is_noneee = purrr::map_lgl(combined_text, ~ match_any(.x, general_nonEEE_patterns))
  )

# Summary
thesis_data |>
  dplyr::summarise(
    total                = dplyr::n(),
    eee_2plus_clean      = sum(n_eee_kw >= 2 & !is_geo & !is_envsoc & !is_noneee),
    eee_1_only           = sum(n_eee_kw == 1),
    eee_0                = sum(n_eee_kw == 0),
    geology_neg          = sum(is_geo & n_eee_kw == 0),
    envsoc_neg           = sum(is_envsoc & n_eee_kw == 0),
    general_neg          = sum(is_noneee & n_eee_kw == 0),
    ambiguous_eee_geo    = sum(n_eee_kw >= 1 & is_geo),
    ambiguous_eee_soc    = sum(n_eee_kw >= 1 & is_envsoc),
    ambiguous_eee_noneee = sum(n_eee_kw >= 1 & is_noneee)
  )
```

## Review Keyword Labels

Before proceeding, it is worth examining a sample of the keyword-labeled theses to
verify the labels are sensible. These samples can be exported to CSV for manual review
and any corrections noted.

```{r}
#| label: review-eee-sample

# Sample of 2+ keyword EEE candidates
thesis_data |>
  dplyr::filter(n_eee_kw >= 2, !is_geo, !is_envsoc, !is_noneee) |>
  dplyr::select(institution, title, n_eee_kw) |>
  dplyr::slice_sample(n = 30) |>
  dplyr::arrange(institution, title) |>
  print(n = 30)
```

```{r}
#| label: review-geo-sample

# Sample of geology negatives
thesis_data |>
  dplyr::filter(is_geo, n_eee_kw == 0) |>
  dplyr::select(institution, title) |>
  dplyr::slice_sample(n = 20) |>
  print(n = 20)
```

```{r}
#| label: export-review-samples
#| eval: true

# Export samples for manual review if desired
thesis_data |>
  dplyr::filter(n_eee_kw >= 2, !is_geo, !is_envsoc, !is_noneee) |>
  dplyr::slice_sample(n = 100) |>
  dplyr::select(institution, title, abstract, n_eee_kw) |>
  readr::write_csv(
    here::here("data", "processed_data", "comparator-theses",
               "training-data", "v2_eee_keyword_sample_for_review.csv")
  )
```

---

# Build Initial Training Set

## Assign Labels

Theses are assigned to one of four groups:

- **EEE positive** (confident): 2+ EEE keyword matches, no geology/env-soc flags
- **Clear negative**: 0 EEE keyword matches, no geology/env-soc flags
- **Near-neighbour negative**: geology or env-soc flag, 0 EEE keywords
- **Ambiguous**: exactly 1 EEE keyword match (used only in semi-supervised rounds)

```{r}
#| label: assign-labels

thesis_data <- thesis_data |>
  dplyr::mutate(
    label_status = dplyr::case_when(
      n_eee_kw >= 2 & !is_geo & !is_envsoc & !is_noneee ~ "positive_eee",
      is_noneee                                           ~ "negative_general",
      n_eee_kw == 0 & (is_geo | is_envsoc)               ~ "negative_near",
      n_eee_kw == 0 & !is_geo & !is_envsoc               ~ "negative_clear",
      TRUE                                                ~ "ambiguous"   # 1 kw, or 2+ kw with geo/soc flag
    ),
    category = dplyr::case_when(
      label_status == "positive_eee"                     ~ "EEE",
      stringr::str_starts(label_status, "negative")      ~ "other",
      TRUE                                               ~ NA_character_
    ),
    category = factor(category, levels = c("EEE", "other"))
  )

thesis_data |> dplyr::count(label_status, category)
```

## Sample Training Set

We downsample to a balanced training set. The target sizes (350 EEE, 200 clear-other,
100 near-neighbour-other, 50 general-other) were chosen based on piloting: going below
~200 per class produces unstable results; going much higher than ~500 per class yields
diminishing returns with this model type. Total negatives (350) match total positives.

```{r}
#| label: build-training-set

set.seed(42)

eee_sample <- thesis_data |>
  dplyr::filter(label_status == "positive_eee") |>
  dplyr::slice_sample(n = 350)

clear_neg_sample <- thesis_data |>
  dplyr::filter(label_status == "negative_clear") |>
  dplyr::slice_sample(n = 200)

near_neg_sample <- thesis_data |>
  dplyr::filter(label_status == "negative_near") |>
  dplyr::slice_sample(n = 100)

general_neg_sample <- thesis_data |>
  dplyr::filter(label_status == "negative_general") |>
  dplyr::slice_sample(n = 50)

training_data <- dplyr::bind_rows(
  eee_sample, clear_neg_sample, near_neg_sample, general_neg_sample
) |>
  dplyr::select(institution, combined_text, category, title, abstract) |>
  dplyr::slice_sample(prop = 1)   # shuffle

cat("Training set composition:\n")
training_data |> dplyr::count(category)
cat("Institution breakdown:\n")
training_data |> dplyr::count(institution)
```

---

# Initial Model Training

## Train/Test Split

```{r}
#| label: train-test-split

set.seed(42)
data_split <- rsample::initial_split(training_data, prop = 0.80, strata = category)
train_data <- rsample::training(data_split)
test_data  <- rsample::testing(data_split)

cv_folds <- rsample::vfold_cv(train_data, v = 5, strata = category)
```

## Text Preprocessing Recipe

The recipe is identical to v1 except the input column name is `combined_text` (consistent
with the pre-processing pipeline in `06_clean_thesis_data.R`).

```{r}
#| label: create-recipe

text_recipe <- recipes::recipe(category ~ combined_text, data = train_data) |>
  textrecipes::step_tokenize(combined_text) |>
  textrecipes::step_stopwords(
    combined_text,
    custom_stopword_source = stopwords::stopwords("en")
  ) |>
  textrecipes::step_tokenfilter(combined_text, min_times = 5, max_tokens = 5000) |>
  textrecipes::step_tfidf(combined_text) |>
  recipes::step_normalize(recipes::all_predictors())
```

## Model Specification and Workflow

```{r}
#| label: model-and-workflow

logistic_spec <- parsnip::logistic_reg(
  penalty  = tune::tune(),
  mixture  = tune::tune()
) |>
  parsnip::set_engine("glmnet") |>
  parsnip::set_mode("classification")

thesis_workflow <- workflows::workflow() |>
  workflows::add_recipe(text_recipe) |>
  workflows::add_model(logistic_spec)
```

## Hyperparameter Tuning

```{r}
#| label: tune-model

tune_grid <- dials::grid_regular(
  dials::penalty(range = c(-5, 0)),
  dials::mixture(range = c(0, 1)),
  levels = 10
)

future::plan(multisession, workers = parallel::detectCores() - 1)

tune_results <- thesis_workflow |>
  tune::tune_grid(
    resamples = cv_folds,
    grid      = tune_grid,
    metrics   = yardstick::metric_set(accuracy, roc_auc, sens, spec),
    control   = tune::control_grid(save_pred = TRUE)
  )

future::plan(sequential)
```

## Select Best Hyperparameters

```{r}
#| label: select-best

best_params <- tune_results |>
  tune::select_best(metric = "roc_auc")

knitr::kable(best_params, caption = "Best hyperparameters (initial model)")
```

## Train Final Initial Model

```{r}
#| label: train-initial-model

final_workflow_v1 <- thesis_workflow |>
  tune::finalize_workflow(best_params)

final_fit_v1 <- final_workflow_v1 |>
  parsnip::fit(data = training_data)
```

---

# Initial Model Evaluation

## Test Set Performance

```{r}
#| label: initial-test-predictions

test_predictions <- final_fit_v1 |>
  predict(test_data) |>
  dplyr::bind_cols(
    final_fit_v1 |> predict(test_data, type = "prob")
  ) |>
  dplyr::bind_cols(test_data)
```

```{r}
#| label: initial-metrics
#| echo: true

test_metrics <- test_predictions |>
  yardstick::metrics(truth = category, estimate = .pred_class)

knitr::kable(test_metrics, digits = 3, caption = "Initial model — test set performance")
```

```{r}
#| label: initial-confusion
#| echo: true

test_predictions |>
  yardstick::conf_mat(truth = category, estimate = .pred_class)
```

**Note on test performance**: High test accuracy on this keyword-labeled test set should
be interpreted carefully. The test set shares the same labeling mechanism as the training
set (keyword rules), so perfect separation is expected where the EEE signal is clear.
The semi-supervised rounds and manual review steps below provide a more meaningful
assessment of generalization.

## Top Predictive Terms

Inspecting the model coefficients is a useful sanity check. EEE-predicting terms should
be biological; Other-predicting terms should reflect near-neighbour fields.

```{r}
#| label: top-terms

top_terms <- final_fit_v1 |>
  workflows::extract_fit_parsnip() |>
  vip::vi(lambda = best_params$penalty) |>
  dplyr::mutate(
    # In glmnet binary classification, Sign = "NEG" predicts the first factor level (EEE)
    Direction    = dplyr::if_else(Sign == "NEG", "EEE", "Other"),
    abs_importance = abs(Importance)
  ) |>
  dplyr::group_by(Direction) |>
  dplyr::slice_max(abs_importance, n = 20) |>
  dplyr::ungroup()

# EEE-predicting terms
top_terms |>
  dplyr::filter(Direction == "EEE") |>
  dplyr::select(Variable, Importance) |>
  print(n = 20)

# Other-predicting terms
top_terms |>
  dplyr::filter(Direction == "Other") |>
  dplyr::select(Variable, Importance) |>
  print(n = 20)
```

---

# Semi-Supervised Refinement

## Strategy for Ambiguous Theses

Theses with exactly 1 EEE keyword match are genuinely ambiguous: some are true EEE
(e.g., a thesis about "habitat restoration" that also discusses nutrient dynamics —
but only the word "habitat" triggered the keyword match). Others are false positives
(e.g., a cancer biology thesis that mentions "habitat" metaphorically).

We apply the trained model to this pool. High-confidence predictions (≥0.80 for either
class) are candidates for inclusion in the next round's training data. A researcher
should review these predictions before including them — the export step below
facilitates this.

```{r}
#| label: get-ambiguous

unlabeled_data <- thesis_data |>
  dplyr::filter(label_status == "ambiguous") |>
  dplyr::select(institution, combined_text, title, abstract, n_eee_kw, is_geo, is_envsoc)

cat("Ambiguous theses (1 keyword match):", nrow(unlabeled_data), "\n")
```

## Round 1: Score Ambiguous Theses

```{r}
#| label: round1-score

unlabeled_predictions_r1 <- final_fit_v1 |>
  predict(unlabeled_data, type = "prob") |>
  dplyr::bind_cols(
    final_fit_v1 |> predict(unlabeled_data, type = "class")
  ) |>
  dplyr::bind_cols(unlabeled_data)

# Summary of predictions
unlabeled_predictions_r1 |>
  dplyr::summarise(
    n_total         = dplyr::n(),
    high_conf_EEE   = sum(.pred_EEE >= 0.80),
    high_conf_other = sum(.pred_other >= 0.80),
    uncertain       = sum(.pred_EEE > 0.20 & .pred_EEE < 0.80)
  )
```

## Round 1: Export High-Confidence Predictions for Review

```{r}
#| label: round1-export
#| eval: false

# Export high-confidence EEE predictions for manual review
unlabeled_predictions_r1 |>
  dplyr::filter(.pred_EEE >= 0.80) |>
  dplyr::arrange(dplyr::desc(.pred_EEE)) |>
  dplyr::select(institution, title, abstract,
                prob_EEE = .pred_EEE, predicted_class = .pred_class,
                is_geo, is_envsoc) |>
  dplyr::mutate(prob_EEE = round(prob_EEE, 3)) |>
  readr::write_csv(
    here::here("data", "processed_data", "comparator-theses",
               "training-data", "v2_round1_high_conf_EEE_for_review.csv")
  )

# Also export uncertain cases for spot-checking
unlabeled_predictions_r1 |>
  dplyr::filter(.pred_EEE > 0.20, .pred_EEE < 0.80) |>
  dplyr::arrange(.pred_EEE) |>
  dplyr::select(institution, title, abstract,
                prob_EEE = .pred_EEE, is_geo, is_envsoc) |>
  dplyr::mutate(prob_EEE = round(prob_EEE, 3)) |>
  readr::write_csv(
    here::here("data", "processed_data", "comparator-theses",
               "training-data", "v2_round1_uncertain_for_review.csv")
  )
```

## Round 1: Incorporate Verified Labels

After manual review, save a CSV with columns `title` and `verified_category` ("EEE" or
"other"). Load it here. **The three chunks below (`round1-load-labels`, `round1-retrain`,
`round1-fit`) are marked `eval: false` because they depend on the manually reviewed file.
Run them interactively once the review CSV exists.**

```{r}
#| label: round1-load-labels
#| eval: false

# After manual review, load the verified labels:
manual_labels_r1 <- readr::read_csv(
  here::here("data", "processed_data", "comparator-theses",
             "training-data", "v2_manually_reviewed_labels_round1.csv")
) |>
  dplyr::select(title, verified_category)

table(manual_labels_r1$verified_category)
```

## Round 1: Retrain

```{r}
#| label: round1-retrain
#| eval: false

# Merge verified labels back into ambiguous pool
newly_labeled_r1 <- unlabeled_data |>
  dplyr::left_join(manual_labels_r1, by = "title") |>
  dplyr::filter(!is.na(verified_category)) |>
  dplyr::mutate(
    category = factor(verified_category, levels = c("EEE", "other"))
  ) |>
  dplyr::select(institution, combined_text, category, title, abstract)

# Expand training set
training_data_v2 <- dplyr::bind_rows(training_data, newly_labeled_r1) |>
  dplyr::slice_sample(prop = 1)

cat("Training set expanded:", nrow(training_data), "→", nrow(training_data_v2), "\n")
training_data_v2 |> dplyr::count(category)
```

```{r}
#| label: round1-fit
#| eval: false

set.seed(1234)

data_split_v2    <- rsample::initial_split(training_data_v2, prop = 0.80, strata = category)
train_data_v2    <- rsample::training(data_split_v2)
test_data_v2     <- rsample::testing(data_split_v2)
cv_folds_v2      <- rsample::vfold_cv(train_data_v2, v = 5, strata = category)

text_recipe_v2 <- recipes::recipe(category ~ combined_text, data = train_data_v2) |>
  textrecipes::step_tokenize(combined_text) |>
  textrecipes::step_stopwords(combined_text,
    custom_stopword_source = stopwords::stopwords("en")) |>
  textrecipes::step_tokenfilter(combined_text, min_times = 5, max_tokens = 5000) |>
  textrecipes::step_tfidf(combined_text) |>
  recipes::step_normalize(recipes::all_predictors())

thesis_workflow_v2 <- workflows::workflow() |>
  workflows::add_recipe(text_recipe_v2) |>
  workflows::add_model(logistic_spec)

future::plan(multisession, workers = parallel::detectCores() - 1)

tune_results_v2 <- thesis_workflow_v2 |>
  tune::tune_grid(
    resamples = cv_folds_v2,
    grid      = tune_grid,
    metrics   = yardstick::metric_set(accuracy, roc_auc, sens, spec),
    control   = tune::control_grid(save_pred = TRUE)
  )

future::plan(sequential)

best_params_v2 <- tune_results_v2 |> tune::select_best(metric = "roc_auc")
final_workflow_v2 <- thesis_workflow_v2 |> tune::finalize_workflow(best_params_v2)
final_fit_v2 <- final_workflow_v2 |> parsnip::fit(data = training_data_v2)
```

```{r}
#| label: round1-evaluate
#| echo: false
#| eval: false

test_preds_v2 <- final_fit_v2 |>
  predict(test_data_v2) |>
  dplyr::bind_cols(final_fit_v2 |> predict(test_data_v2, type = "prob")) |>
  dplyr::bind_cols(test_data_v2)

knitr::kable(
  test_preds_v2 |> yardstick::metrics(truth = category, estimate = .pred_class),
  digits = 3, caption = "Round 1 model — test set performance"
)

test_preds_v2 |> yardstick::conf_mat(truth = category, estimate = .pred_class)
```

## Round 2 (and Further Rounds)

**DIDN'T NEED THIS STEP BECAUSE I CLASSIFIED ALL UNCERTAIN THESES IN THE FIRST ROUND**

Repeat the score → review → retrain cycle for the still-uncertain theses until the
model stabilizes or the remaining uncertain pool is small enough to review exhaustively.
The code below provides the template; duplicate and increment version numbers as needed.
**Run these chunks interactively after completing Round 1 manual review.**

```{r}
#| label: round2-score
#| eval: false
#| echo: false

# Score the remaining uncertain theses with the round 1 model,
# excluding any theses already manually reviewed in round 1
still_uncertain <- unlabeled_predictions_r1 |>
  dplyr::filter(.pred_EEE > 0.20, .pred_EEE < 0.80) |>
  dplyr::anti_join(manual_labels_r1, by = "title") |>
  dplyr::select(institution, combined_text, title, abstract, n_eee_kw, is_geo, is_envsoc)

cat("Still uncertain after round 1:", nrow(still_uncertain), "\n")

r2_scores <- final_fit_v2 |>
  predict(still_uncertain, type = "prob") |>
  dplyr::bind_cols(final_fit_v2 |> predict(still_uncertain, type = "class")) |>
  dplyr::bind_cols(still_uncertain)

r2_scores |>
  dplyr::summarise(
    high_conf_EEE   = sum(.pred_EEE >= 0.80),
    high_conf_other = sum(.pred_other >= 0.80),
    still_uncertain = sum(.pred_EEE > 0.20 & .pred_EEE < 0.80)
  )
```

```{r}
#| label: round2-export
#| eval: false
#| echo: false

r2_scores |>
  dplyr::filter(.pred_EEE >= 0.80 | .pred_other >= 0.80) |>
  dplyr::arrange(dplyr::desc(.pred_EEE)) |>
  dplyr::select(institution, title, abstract,
                prob_EEE = .pred_EEE, predicted_class = .pred_class) |>
  dplyr::mutate(prob_EEE = round(prob_EEE, 3)) |>
  readr::write_csv(
    here::here("data", "processed_data", "comparator-theses",
               "training-data", "v2_round2_high_conf_for_review.csv")
  )

r2_scores |>
  dplyr::filter(.pred_EEE > 0.20, .pred_EEE < 0.80) |>
  dplyr::arrange(.pred_EEE) |>
  dplyr::select(institution, title, abstract,
                prob_EEE = .pred_EEE, is_geo, is_envsoc) |>
  dplyr::mutate(prob_EEE = round(prob_EEE, 3)) |>
  readr::write_csv(
    here::here("data", "processed_data", "comparator-theses",
               "training-data", "v2_round2_uncertain_for_review.csv")
  )
```

---

# Final Model and Evaluation

Once iterative refinement is complete, retrain on the full labeled set using the code
pattern above. The chunk below assumes the final model is `final_fit_v2` (or the last
iteration's model); update the object name as appropriate. **Run interactively after
all review rounds are complete.**

```{r}
#| label: final-performance
#| eval: false

# Replace final_fit_v2 with the last iteration's fit object when ready
final_model <- final_fit_v2

test_predictions_final <- final_model |>
  predict(test_data_v2) |>
  dplyr::bind_cols(final_model |> predict(test_data_v2, type = "prob")) |>
  dplyr::bind_cols(test_data_v2)

final_metrics <- test_predictions_final |>
  yardstick::metrics(truth = category, estimate = .pred_class)

sensitivity_final <- test_predictions_final |>
  yardstick::sens(truth = category, estimate = .pred_class) |>
  dplyr::pull(.estimate)

specificity_final <- test_predictions_final |>
  yardstick::spec(truth = category, estimate = .pred_class) |>
  dplyr::pull(.estimate)
```

```{r}
#| label: final-metrics-table
#| echo: false
#| eval: false

knitr::kable(final_metrics, digits = 3, caption = "Final model — test set performance")
```

- **Sensitivity (EEE recall)**: (run `final-performance` chunk to populate)
- **Specificity (Other recall)**: (run `final-performance` chunk to populate)

```{r}
#| label: final-confusion
#| echo: false
#| eval: false

test_predictions_final |>
  yardstick::conf_mat(truth = category, estimate = .pred_class)
```

---

# Save Final Model

```{r}
#| label: save-model
#| eval: false

saveRDS(
  final_model,
  here::here("data", "processed_data", "comparator-theses",
             "training-data", "eee_text_classifier_v2.rds")
)

model_info_v2 <- list(
  model_file     = "eee_text_classifier_v2.rds",
  training_date  = Sys.Date(),
  training_size  = nrow(training_data_v2),
  n_eee          = sum(training_data_v2$category == "EEE"),
  n_other        = sum(training_data_v2$category == "other"),
  test_accuracy  = final_metrics |>
    dplyr::filter(.metric == "accuracy") |>
    dplyr::pull(.estimate),
  sensitivity    = sensitivity_final,
  specificity    = specificity_final,
  approach       = "Multi-institutional keyword-seeded labels with near-neighbour negatives",
  institutions   = sort(unique(thesis_data$institution)),
  features       = "title + abstract (combined_text)"
)

saveRDS(
  model_info_v2,
  here::here("data", "processed_data", "comparator-theses",
             "training-data", "eee_text_classifier_v2_model_info.rds")
)
```

---

# Applying the Model to New Data

Once the model is saved, applying it to a new batch of thesis metadata requires three
steps: load the model, prepare the input in the expected format, and call `predict()`.

**Input requirements**: the new data must contain `title` and `abstract` columns. The
`combined_text` feature used by the model is constructed by concatenating these two
fields, exactly as done during training.

```{r}
#| label: apply-to-new-data
#| eval: false

# 1. Load the saved model
final_model <- readRDS(
  here::here("data", "processed_data", "comparator-theses",
             "training-data", "eee_text_classifier_v2.rds")
)

# 2. Load and prepare new thesis metadata
#    Required columns: title, abstract (plus any ID columns you want to carry through)
new_theses <- readr::read_csv("path/to/new_theses.csv") |>
  dplyr::mutate(
    abstract      = dplyr::if_else(is.na(abstract), "", abstract),
    combined_text = paste(title, abstract, sep = " ")
  )

# 3. Generate predictions
#    .pred_class  : "EEE" or "other"
#    .pred_EEE    : probability of EEE class (0–1)
predictions <- final_model |>
  predict(new_theses) |>
  dplyr::bind_cols(final_model |> predict(new_theses, type = "prob")) |>
  dplyr::bind_cols(new_theses)

# 4. Inspect and filter
#    Use .pred_class for a hard classification, or threshold .pred_EEE
#    directly if you need finer control (e.g. ≥ 0.70 for higher recall)
eee_theses <- predictions |>
  dplyr::filter(.pred_class == "EEE")

cat(sprintf("Classified %d / %d theses as EEE\n", nrow(eee_theses), nrow(new_theses)))
```

The `Category` column written by `07_classify_theses.R` is produced by exactly this
workflow: it loads the `.rds` model, constructs `combined_text`, and appends the
predicted class to the cleaned thesis CSV for each institution.

---

# Appendix: Top Predictive Terms (Final Model)

```{r}
#| label: top-terms-final
#| eval: false

top_terms_final <- final_model |>
  workflows::extract_fit_parsnip() |>
  vip::vi(lambda = best_params_v2$penalty) |>
  dplyr::mutate(
    Direction      = dplyr::if_else(Sign == "NEG", "EEE", "Other"),
    abs_importance = abs(Importance)
  ) |>
  dplyr::group_by(Direction) |>
  dplyr::slice_max(abs_importance, n = 20) |>
  dplyr::ungroup()

top_terms_final |>
  ggplot2::ggplot(ggplot2::aes(
    x = Importance, y = reorder(Variable, Importance), fill = Direction
  )) +
  ggplot2::geom_col() +
  ggplot2::facet_wrap(~Direction, scales = "free_y") +
  ggplot2::labs(
    title = "Top 20 Predictive Terms by Category (Final Model v2)",
    y = NULL, x = "Coefficient"
  ) +
  ggplot2::theme_minimal()

ggplot2::ggsave(
  here::here("data", "processed_data", "comparator-theses",
             "training-data", "v2_top_predictive_terms.png"),
  width = 12, height = 8
)
```

---

## Session Info

```{r}
#| label: session-info

sessionInfo()
```
