---
title: "PRISMA-Style Pipeline Flow Diagram"
subtitle: "Thesis metadata from raw collection through classifier training and application — LDP comparator identification study"
author: "Jason Pither"
date: today
format:
  html:
    self-contained: true
    toc: false
execute:
  echo: false
  warning: false
  message: false
---

## Purpose

This document renders a PRISMA-style flow diagram tracing thesis metadata from
initial scraping across eight Canadian institutions through cleaning, keyword-seeded
label assignment, classifier training and semi-supervised refinement, and final
application to produce EEE / other predictions for all 16,900 clean records.

### Data sources for all counts

| Count | Source |
|---|---|
| Raw thesis counts (Stage 1) | Files in `data/processed_data/comparator-theses/raw/` |
| Clean thesis counts (Stage 2) | Files in `data/processed_data/comparator-theses/clean/` |
| Training set composition | `thesis_classification_model_training.qmd`; `eee_text_classifier_v2_model_info.rds` |
| Semi-supervised review counts | `v2_manually_reviewed_labels_round1.csv` |
| Final classified counts | Files in `data/processed_data/comparator-theses/classified/` |

### Pipeline stage descriptions

**Stage 1 — Thesis metadata collection (scripts 01-1 through 01-6).**
Thesis metadata (title, abstract, author, year, degree, institution) is scraped from
institutional repositories and merged. Sources: UBC cIRcle, University of Alberta and
University of Toronto Scholaris repositories, and McGill eScholarship. Records for
Manitoba, Guelph, Regina, and WLU were retrieved from Library and Archives Canada (LAC).

**Stage 2 — Cleaning (`02_clean_theses.R`).**
Raw CSVs are standardised to a common schema, filtered to LDP-relevant institutions
and degree types (masters and doctoral), and written to `clean/` as one file per
institution.

**Stage 3 — Keyword-seeded label assignment.**
Each clean thesis is assigned a provisional label using keyword matching on combined
title + abstract text. Positive EEE labels require ≥ 2 EEE keyword matches with no
near-neighbour flags. Clear negatives are records with no keyword matches;
near-neighbour negatives are records matching geology or environmental-social keywords.
Ambiguous records match exactly 1 EEE keyword.

**Stage 4 — Initial model training (v1).**
A tidymodels TF-IDF elastic-net logistic regression classifier is trained on the
keyword-labeled training set (350 EEE + 350 other = 700 records total), using 5-fold
cross-validation. The initial model is applied to ambiguous records to generate
predicted EEE probabilities.

**Stage 5 — Semi-supervised Round 1.**
Ambiguous records with high model confidence or uncertainty are exported for manual
review (442 records). Manually verified labels (235 EEE, 207 other) are added to the
training set, and the final v2 model is retrained on the expanded set.

**Stage 6 — Final model (v2) and classifier application (`03_apply_classifier.R`).**
The v2 model (`eee_text_classifier_v2.rds`; 585 EEE + 557 other = 1,142 records) is
applied to all 16,900 clean theses, appending `Category` (EEE / other) and `prob_EEE`
columns to each institution's CSV.

---

```{r}
#| label: prisma-diagram
#| cache: false
#| fig-cap: >
#|   PRISMA-style flow of thesis metadata from raw collection through classifier
#|   training and application to the final EEE / other predictions.
#|   Blue boxes = main data flow; light yellow boxes = processing / model steps;
#|   amber box = ambiguous / held-for-review records; light green boxes = positive
#|   or verified-label records; dark green box = final output;
#|   red boxes = excluded or not-used records.

library(DiagrammeR)

dot_src <- '
digraph prisma {

  graph [rankdir  = TB,
         splines  = ortho,
         nodesep  = 0.55,
         ranksep  = 0.60,
         fontname = "Helvetica",
         bgcolor  = "white"]

  node [fontname = "Helvetica",
        fontsize  = 9,
        style     = "filled,rounded",
        shape     = rectangle,
        margin    = "0.18,0.12",
        penwidth  = 1.2]

  edge [arrowsize = 0.7,
        penwidth  = 1.0]

  //=================================================================
  // STAGE 1 — RAW COLLECTION
  //=================================================================

  raw [label    = "THESIS METADATA — RAW COLLECTION\nScripts 01-1 through 01-6\n8 institutions (UBC, Alberta, Toronto, McGill,\n Manitoba, Guelph, Regina, WLU)\nn = 19,641 records",
       fillcolor = "#D6EAF8", color = "#1A5276", width = 4.2,
       fontname  = "Helvetica-Bold"]

  //=================================================================
  // STAGE 2 — CLEANING
  //=================================================================

  clean_step [label = "02_clean_theses.R\nStandardise schema; filter to LDP-relevant\ninstitutions and degree types (masters + doctoral)",
              fillcolor = "#FEF9E7", color = "#7D6608", width = 4.2,
              style = "filled,rounded"]

  clean [label    = "CLEAN THESES\n8 institutions\nn = 16,900 records",
         fillcolor = "#D6EAF8", color = "#1A5276", width = 4.2]

  e_clean [label = "Excluded at cleaning\nn = 2,741\n(non-LDP institutions; ineligible degree types;\nmissing / malformed records)",
           fillcolor = "#FADBD8", color = "#922B21", width = 3.0]

  //=================================================================
  // STAGE 3 — KEYWORD-SEEDED LABEL ASSIGNMENT
  //=================================================================

  kw_step [label = "Keyword-seeded label assignment\n(thesis_classification_model_training.qmd)",
           fillcolor = "#FEF9E7", color = "#7D6608", width = 4.2,
           style = "filled,rounded"]

  kw_pos   [label = "Positive EEE labels\n(≥ 2 EEE keyword matches,\nno near-neighbour flags)\nn = 350",
            fillcolor = "#D5F5E3", color = "#1D6A39", width = 3.0]

  kw_neg   [label = "Negative labels for training\nClear negatives: n = 200\nNear-neighbour (geology/env-soc): n = 100\nGeneral negatives: n = 50\nTotal: n = 350",
            fillcolor = "#FADBD8", color = "#922B21", width = 3.0]

  kw_ambig [label = "Ambiguous\n(exactly 1 EEE keyword match)\nHeld for semi-supervised review",
            fillcolor = "#FFF3CD", color = "#7D6608", width = 3.0]

  //=================================================================
  // STAGE 4 — INITIAL MODEL TRAINING (v1)
  //=================================================================

  train_v1 [label = "Initial model training\nTF-IDF + elastic-net logistic regression\n5-fold CV  |  training set n = 700\n(350 EEE + 350 other)",
            fillcolor = "#FEF9E7", color = "#7D6608", width = 4.2,
            style = "filled,rounded"]

  //=================================================================
  // STAGE 5 — SEMI-SUPERVISED ROUND 1
  //=================================================================

  semisup [label = "Semi-supervised Round 1\nv1 model applied to ambiguous records;\nhigh-confidence and uncertain subsets exported\nfor manual review (n = 442 reviewed)",
           fillcolor = "#FEF9E7", color = "#7D6608", width = 4.2,
           style = "filled,rounded"]

  review  [label = "Manual review output\nVerified EEE:   n = 235\nVerified other: n = 207\nTotal added:    n = 442",
           fillcolor = "#D5F5E3", color = "#1D6A39", width = 3.0]

  //=================================================================
  // STAGE 6 — FINAL MODEL (v2) + APPLICATION
  //=================================================================

  train_v2 [label = "Refined model training\nExpanded training set n = 1,142\n(585 EEE + 557 other)\nTest accuracy = 99.6%  |  Sensitivity = 99.1%  |  Specificity = 100%",
            fillcolor = "#FEF9E7", color = "#7D6608", width = 4.2,
            style = "filled,rounded"]

  apply [label = "03_apply_classifier.R\nv2 model applied to all 16,900 clean theses",
         fillcolor = "#FEF9E7", color = "#7D6608", width = 4.2,
         style = "filled,rounded"]

  //=================================================================
  // FINAL OUTPUT
  //=================================================================

  final [label = "CLASSIFIED THESES — 8 institutions  |  n = 16,900\nEEE: n = 1,682     Other: n = 15,218",
         fillcolor = "#82E0AA", color = "#1D6A39", width = 4.2,
         style = "filled,bold,rounded", fontsize = 11]

  //=================================================================
  // MAIN FLOW EDGES
  //=================================================================

  raw      -> clean_step
  clean_step -> clean
  clean    -> kw_step
  kw_step  -> train_v1
  train_v1 -> semisup
  semisup  -> train_v2
  train_v2 -> apply
  apply    -> final

  //=================================================================
  // SIDE / DETAIL EDGES  (constraint = false: no rank effect)
  //=================================================================

  clean_step -> e_clean  [constraint = false, style = dashed, color = "#A93226"]
  kw_step    -> kw_pos   [constraint = false, style = dashed, color = "#1D6A39"]
  kw_step    -> kw_neg   [constraint = false, style = dashed, color = "#A93226"]
  kw_step    -> kw_ambig [constraint = false, style = dashed, color = "#7D6608"]
  semisup    -> review   [constraint = false, style = dashed, color = "#1D6A39"]

  //=================================================================
  // RANK GROUPS  (force same horizontal level)
  //=================================================================

  { rank = same;  clean_step;  e_clean  }
  { rank = same;  kw_step;     kw_pos   }
  { rank = same;  train_v1;    kw_neg   }
  { rank = same;  semisup;     kw_ambig }
  { rank = same;  train_v2;    review   }

  //=================================================================
  // INVISIBLE ORDERING EDGES  (enforce left-to-right within ranks)
  //=================================================================

  clean_step -> e_clean  [style = invis]
  kw_step    -> kw_pos   [style = invis]
  train_v1   -> kw_neg   [style = invis]
  semisup    -> kw_ambig [style = invis]
  train_v2   -> review   [style = invis]

}
'

grViz(dot_src, height = 900, width = "100%")
```

---

```{r}
#| label: export-svg-pdf
#| cache: false

library(DiagrammeRsvg)
library(rsvg)
library(here)

svg_str  <- DiagrammeRsvg::export_svg(grViz(dot_src))
svg_path <- here::here("scripts", "thesis_classification_prisma.svg")
pdf_path <- here::here("scripts", "thesis_classification_prisma.pdf")

# Write SVG
writeLines(svg_str, svg_path)

# Convert SVG file to PDF
rsvg::rsvg_pdf(svg = svg_path, file = pdf_path)

message("Exported:\n  ", svg_path, "\n  ", pdf_path)
```
